---
title: "Intro to MAF"
author: Nicole Black
date: 07/31/22
---

###### (1) Set Up

```{r setup}
    knitr::opts_knit$set(root.dir = normalizePath("/Users/amirarya/Desktop/qbio_490_amir/week6_maf copy")) 
```

```{r}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager :: install("maftools")
library(maftools)
library(TCGAbiolinks)
```

Before we get into MAF, we need to read in the clinical data (created in week 4) so it can be merged with our MAF data later on.

```{r Exercise 1.1}
clinical <- read.csv("/Users/amirarya/Desktop/qbio_490_amir/brca_clinical_data.csv")
```

We also need to change the bcr_patient_barcode column name that way the MAF package can read our clinical file.

```{r Exercise 1.2}
colnames(clinical)[ colnames(clinical) == "bcr_patient_barcode" ] <- "Tumor_Sample_Barcode"
```

After changing the column name, re-save the clinical data frame to your computer.

```{r}
# change the file path! make sure it's in your week4 folder
write.csv(clinical, "/Users/amirarya/Desktop/qbio_490_amir/brca_clinical_data.csv", row.names = FALSE)
```

###### (2) MAF Package

Now, we can query in the MAF files using GDCquery()/GDCdownload()/GDCprepare() and read.maf(). 

```{r Exercise 2.1}
maf_query <- GDCquery(
    project = "TCGA-BRCA", 
    data.category = "Simple Nucleotide Variation", 
    access = "open", # we only have access to somatic mutations which are open access
    data.type = "Masked Somatic Mutation", 
    workflow.type = "Aliquot Ensemble Somatic Variant Merging and Masking"
)
GDCdownload(maf_query)
maf <- GDCprepare(maf_query) # as long as it runs, ignore any errors

maf_object <- read.maf(maf = maf, 
                       clinicalData = clinical,
                       isTCGA = TRUE)
```

###### (3) Exploring the MAF Object

Call str() on maf_object.

```{r Exercise 3.1}
str(maf_object)
```

Notice the @ and $ signs. What do each of them represent?

  @ represents: accessing a data frame
  $ represents: acessing a collum
  
List all of the data frames within the maf_object by typing maf_object@ then either waiting for a popup, or if no popup appears, by hitting tab.

```{r Exercise 3.2}
maf_object@data$X1
```

Call head() and str() on the @data df and the @clinical.data df. These will be the main data frames we use in our MAF analysis.

```{r Exercise 3.3}
head(maf_object@data)
str(maf_object@data)
```

View the first 110 items in Tumor_Sample_Barcode from @data and @clinical.data.

```{r Exercise 3.4}
maf_object@data$Tumor_Sample_Barcode[1:110]
maf_object@clinical.data$Tumor_Sample_Barcode[1:110]

```

What do you notice about these two lists? one of them is all of the same- showing pateint can show up mutliple times - has 108 different mutations 

View the Tumor_Sample_Barcode column and the HGVSp_Short column for all TP53 mutations (i.e. Hugo_Symbol == "TP53"). Hint: boolean masking!

```{r Exercise 3.5}
mask1 <- (maf_object@data$Hugo_Symbol == "TP53")
maf_object@data$Tumor_Sample_Barcode[mask1]
maf_object@data$HGVSp_Short[mask1]
```

Call head() on the Amino_acids column in @data.

```{r Exercise 3.6}
# write code here
```

What mutation could cause a S/R change?

What about an A/T change?

Call head() on the HGVSp_Short column in @data.

```{r Exercise 3.7}
# write code here
```

What does p.P728Afs*11 mean? (Do some google-ing!)

```{r Exercise 3.8}
# use this code block for the following questions:
```
  
  How many rows/columns are there in @data?
  
  What type of data is stored in the rows? What about the columns?
  
  How many levels are there in Tumor_Sample_Barcode (from @data)?
  Are there duplicate values in Tumor_Sample_Barcode?
  
  How many unique values are there in Amino_acids?
  Are there duplicate values in Amino_acids?
  
  How many unique values are there in HGVSp_Short?
  Are there duplicate values in HGVSp_Short?
